---
title: "Internship_Task"
author: "Dennis Ndubi"
date: "`r Sys.Date()`"
output: html_document
---
**CEMA Internship Task**

The dataset for this analysis contains monthly data for children <5 years, disaggregated at a county level for the period January 2021 to June 2023.

The dataset contains the following varibales:

- Period (months from January 2021 to June 2023)
- County (the 47 counties in Kenya)
- Total number of children dewormed (Total Dewormed)
-Number of children <5 years with acute malnutrition (Acute Malnutrition)
- Number of children stunted (0-6 months, 6-23 months, 24-59 months)
- Number of children <5 years with diarrhe (Diarrhea cases)
- Number of children who are underweight (0-6 months, 6-23 months, 24-59 months)

**Research Questions**

- What is the prevalence of diarrhea cases across the counties?
- what are the deworming efforts across the counties
- What is the Prevalence of Acute Malnutrition across the counties?
- What is the prevalence of Stunted Growth and underweight Across the counties?
- Is there correlation between Stunted growth and presence of  underweight in children below 5 years.

**Key Finding from the Analysis**

The following are some of the insights gathered from the data set:

* `Nairobi County` is performing well in terms of deworming efforts followed by `Nakuru County`, While `Nyandaru County` is the worst performing.
* In terms of Acute Malnutrition cases, `Wajir county` tops     followed by `Turkana County`. This can be justified by the  prolonged droughts and hunger status experienced in those  areas.`Uasin Gishu County` has  the least cases hence the county  with most health children.
* In terms of disease indication, `Nairobi County` followed by `Turkana County` tops among the counties with highest number of diarrhea cases   which might associated with poor hygiene while `Nyandarua county` has the least amount of diarrhea reported cases.
* There is significant evidence that children who have stunted growth have high probability to  be underweight. Top counties with stunted growth cases are `Nairobi` and `Kilifi` while `Boment` has the least number of stunted growth cases. In terms of Underweight children,`Turkana` and `Nairobi` tops while Still `Bomet` has the least.

*ANALYSIS*

**Importing Necessary Libraries**

```{r echo=TRUE, message=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
library(dplyr)
library(tidyr)
library(skimr)
```

**loading data**
```{r}
data<-read.csv("cema_internship_task_2023.csv", header = TRUE)
```

```{r}
View(data) #Viewing the imported data
```

**Check if Data Imported Correctly**

```{r}
head(data,2) #the  top two observations
```

```{r }
tail(data,2) #bottom 5 observations
```

```{r}
nrow(data)# Number of records in the data set
ncol(data) # Number of columns (attributes) in the data set
dim(data) # shape of our data set
```
```{r}
names(data) # column names of the data set
```

```{r}
d_types=lapply(data,class) #checking datatypes
d_types
```

**Renaming the Variables for clarity and assigning the  correct data types**

```{r }
data<-data %>% rename(Period=period,    
County=county,                 
Total_Dewormed=Total.Dewormed,        
Acute_Malnutrition=Acute.Malnutrition,    
Stunted_6_23_Months=stunted.6.23.months ,     
Stunted_0_6_Months= stunted.0..6.months, 
Stunted_24_59_Months=stunted.24.59.months,    
Diarrhoea_Cases=diarrhoea.cases,
Underweight_0_6_Months=Underweight.0..6.months,
Underweight_6_23_Months=Underweight.6.23.months,
Underweight_24_59_Months=Underweight.24.59.Months)

colnames(data) #renamed columns

```

**Exploring the data set**

```{r}
skim_without_charts(data) #Summary of how the data set looks
```

```{r}
data$Period<-as.Date(paste("01", data$Period, sep = "-"), format = "%d-%b-%Y")
```

```{r}
d_types=lapply(data,class) #rechecking datatypes after typecasting
d_types
```

**Dropping null values from our data set**

```{r}
data=drop_na(data)
```

**Top and Bottom 5 Counties with highest Acute Malnutrition cases**
```{r echo=FALSE}
Counnties_AM_Top<-data%>%
  group_by(County)%>%
  summarise(Agg_Acute_Mal=sum(Acute_Malnutrition))%>%
  arrange(desc(Agg_Acute_Mal))%>%
  select(County,Agg_Acute_Mal)
Counnties_AM_Bottom<-data%>%
  group_by(County)%>%
  summarise(Agg_Acute_Mal=sum(Acute_Malnutrition))%>%
  arrange(Agg_Acute_Mal)%>%
  select(County,Agg_Acute_Mal)
```

```{r}
Top5_C_by_AM<-head(Counnties_AM_Top,5)
Top5_C_by_AM
Bottom5_C_by_AM<-head(Counnties_AM_Bottom,5)
Bottom5_C_by_AM
```


**Top  and Bottom 5 Counties with highest Deworming cases**
```{r echo=FALSE,results='markup'}
Counties_Top<-data%>%
  group_by(County)%>%
  summarise(Agg_T_Dewormed=sum(Total_Dewormed))%>%
  arrange(desc(Agg_T_Dewormed))%>%
  select(County,Agg_T_Dewormed)
Counties_Bottom<-data%>%
  group_by(County)%>%
  summarise(Agg_T_Dewormed=sum(Total_Dewormed))%>%
  arrange(Agg_T_Dewormed)%>%
  select(County,Agg_T_Dewormed)

```
```{r}
Top_5<-head(Counties_Top,5)
Top_5
Bottom_5<-head(Counties_Bottom,5)
Bottom_5
```


**Top 5 Counties with highest Diarrhea Cases**
```{r echo=FALSE}
Counties_Top_D<-data%>%
  group_by(County)%>%
  summarise(Agg_Diarrhea=sum(Diarrhoea_Cases))%>%
  arrange(desc(Agg_Diarrhea))%>%
  select(County,Agg_Diarrhea)
Counties_Bottom_D<-data%>%
  group_by(County)%>%
  summarise(Agg_Diarrhea=sum(Diarrhoea_Cases))%>%
  arrange(Agg_Diarrhea)%>%
  select(County,Agg_Diarrhea)
```

```{r}
Top_5<-head(Counties_Top_D,5)
Top_5
Bottom_5<-head(Counties_Bottom_D,5)
Bottom_5
```

**Top 5 Counties with highest Stunted Growth cases**
```{r echo=FALSE}
Counties_Top_S<-data%>%
  group_by(County)%>%
  mutate(Stunted_Overall=Stunted_0_6_Months+Stunted_6_23_Months+Stunted_24_59_Months)%>%
  summarise(Agg_Stunted_Overall=sum(Stunted_Overall))%>%
  arrange(desc(Agg_Stunted_Overall))%>%
  select(County,Agg_Stunted_Overall)
Counties_Bottom_S<-data%>%
  group_by(County)%>%
  mutate(Stunted_Overall=Stunted_0_6_Months+Stunted_6_23_Months+Stunted_24_59_Months)%>%
  summarise(Agg_Stunted_Overall=sum(Stunted_Overall))%>%
  arrange(Agg_Stunted_Overall)%>%
  select(County,Agg_Stunted_Overall)
```

```{r}
Top_5<-head(Counties_Top_S,5)
Top_5
Bottom_5<-head(Counties_Bottom_S,5)
Bottom_5
```

**Top 5 Counties with highest Underweight cases**
```{r echo=FALSE}
Counties_Top_U<-data%>%
  group_by(County)%>%
  mutate(Underweight_Overall=Underweight_0_6_Months+Underweight_6_23_Months+Underweight_24_59_Months)%>%
  summarise(Agg_Underweight_Overall=sum(Underweight_Overall))%>%
  arrange(desc(Agg_Underweight_Overall))%>%
  select(County,Agg_Underweight_Overall)
Counties_Bottom_U<-data%>%
  group_by(County)%>%
  mutate(Underweight_Overall=Underweight_0_6_Months+Underweight_6_23_Months+Underweight_24_59_Months)%>%
  summarise(Agg_Underweight_Overall=sum(Underweight_Overall))%>%
  arrange(Agg_Underweight_Overall)%>%
  select(County,Agg_Underweight_Overall)
```

```{r}
Top_5<-head(Counties_Top_U,5)
Top_5
Bottom_5<-head(Counties_Bottom_U,5)
Bottom_5
```

**Health Indicators**

Variables taken into consideration are: 
- Total_Dewormed  to indicate deworming efforts 
- Acute_Multnutrition to show child nutrition

```{r }
Summary_dewormed=summary(data$Total_Dewormed)
summary_mulnutrition=summary(data$Acute_Malnutrition)

```

Summary Statistics for Total Dewormed

```{r}
print(Summary_dewormed)
```

Summary Statistics for Acute Malnutrition

```{r}
print(summary_mulnutrition)
```

Dewarming trend over the period

```{r}
data1<-data%>%
  group_by(County,Period)%>%
  summarise(Agg_Total_Dewormed=sum(Total_Dewormed))%>%
  select(Period,Agg_Total_Dewormed)
ggplot(data1, aes(x = Period, y = Agg_Total_Dewormed)) +
  geom_line(color = "blue") +
  labs(title = "Time Series Visualization of Aggregate Total Dewormed",
       x = "Period", y = "Aggregate Total Dewormed")+
  scale_x_date(labels = date_format("%b %Y"))
```

Distribution of Mean Acute Malnutrition Across the counties per Year

```{r}
g_data<- data%>%group_by(Year=year(Period))%>%
  summarize(Mean_Acute_Malnutrition = mean(Acute_Malnutrition))
ggplot(g_data, aes(x = Year, y = Mean_Acute_Malnutrition)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Mean Acute Malnutrition by Year",
       x = "Year", y = "Mean Acute Malnutrition")
```

**Is there relationship between Stunted growth and underweight among children across the counties**

Perform 3 correlations  per age-group
- Stunted O_6 Months VS Overweight 0_6 Months
```{r}
ggplot(data=data,mapping = aes(x=Stunted_0_6_Months, y=Underweight_0_6_Months))+
  geom_point()+
  geom_smooth()+
  labs(x="Stunted between 0-<6 Months", y="Underweight 0-<6 Months")
lm1=lm(data=data,Underweight_0_6_Months~Stunted_0_6_Months)
r_squared <- sprintf("R-squared = %.3f", summary(lm1)$r.squared)
r_squared
 
```

- Stunted 6_23 Months VS Overweight 6_23 Months

```{r}
ggplot(data=data,mapping = aes(x=Stunted_6_23_Months, y=Underweight_6_23_Months))+
  geom_point()+
  geom_smooth()+
  labs(x="Stunted between 6-23 Months", y="Underweight 6-23 Months")
lm2=lm(data=data,Underweight_6_23_Months~Stunted_6_23_Months)
r_squared <- sprintf("R-squared = %.3f", summary(lm2)$r.squared)
r_squared
 
```

- Stunted 24-59 Months Vs Underweight 24-59 Months

```{r}
ggplot(data=data,mapping = aes(x=Stunted_24_59_Months, y=Underweight_24_59_Months))+
  geom_point()+
  geom_smooth()+
  labs(x="Stunted between 24-59 Months", y="Underweight 24-59 Months")
lm3=lm(data=data,Underweight_24_59_Months~Stunted_24_59_Months)
r_squared <- sprintf("R-squared = %.3f", summary(lm3)$r.squared)
r_squared
 
```

Overall Stunted Vs Underweight

```{r}
data<-data%>%
  mutate(Stunted_Overall=Stunted_0_6_Months+Stunted_6_23_Months+Stunted_24_59_Months,Underweight_Overall=Underweight_0_6_Months+Underweight_6_23_Months+Underweight_24_59_Months)
ggplot(data=data,mapping = aes(x=Stunted_Overall, y=Underweight_Overall))+
  geom_point()+
  geom_smooth()+
  labs(x="Stunted_Overall", y="Underweight_Overall")
lm_general=lm(data=data,Underweight_Overall~Stunted_Overall)
r_squared <- sprintf("R-squared = %.3f", summary(lm_general)$r.squared)
r_squared
 
```

Conclusion: There is strong correlation between stunted growth and underweight traits

**Is there relationship between Acute Malnutrition  and Stunted growth/Underweight**

```{r}
ggplot(data=data,mapping = aes(x=Acute_Malnutrition, y=Stunted_Overall))+
  geom_point()+
  geom_smooth()+
  labs(x="Acute_Malnutrition", y="Stunted_Overall")
lm_1=lm(data=data,Stunted_Overall~Acute_Malnutrition)
r_squared <- sprintf("R-squared = %.3f", summary(lm_1)$r.squared)
r_squared
 
```

```{r}
ggplot(data=data,mapping = aes(x=Acute_Malnutrition, y=Underweight_Overall))+
  geom_point()+
  geom_smooth()+
  labs(x="Acute_Malnutrition", y="Underweight_Overall")
lm_g=lm(data=data,Underweight_Overall~Acute_Malnutrition)
r_squared <- sprintf("R-squared = %.3f", summary(lm_g)$r.squared)
r_squared
 
```

**Distribution of diarrhea cases across the counties**

```{r}
data_agg_D <- data %>%
  group_by(County) %>%
  summarise(Agg_Diarrhea_cases = sum(Diarrhoea_Cases))
ggplot(data_agg_D, aes(x = "All Counties", y = Agg_Diarrhea_cases)) +
  geom_boxplot() +
  labs(title = "Boxplot of Diarrhea Cases Across All Counties",
       x = "Counties", y = "Aggregate Diarrhea Cases")
```

